Spatio-angular fluorescence microscopy I. Basic theory (Draft: \today)

### 2018-12-14 papers/basic-theory/paper/paper.tex
Rudolf's edits.

\section{Introduction}
Fluorescence microscopes are widely used in the biological sciences for
measuring ensembles of fluorescent molecules. While most fluorescence microscopy
experiments are designed to measure only the spatial distribution of
fluorophores, a growing number of experiments seek to measure both the spatial
and angular distributions of fluorophores by the use of polarizers
\cite{vrabioiu2006, mattheyses2010, mehta2016, mcquilken2017, zhanghao2017} or
point spread function engineering \cite{agrawal2012, zhang2018}.

@@

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think it accurately describes the
way many models of fluorescence microscopes treat fluorophores, and we use the
term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, monopole absorber/emitters do
not exist in nature. All physical fluorophores absorb and emit radiation with
dipole or higher-order transitions, and these transitions are always oriented in
space. For a classical mental model of fluorophores we imagine each dipole as a
small oriented antenna (with incoherent absorption and emission moments) where
electrons are constrained to move along a single direction. With this model we
can argue classically that monopole radiators do not exist because they would
violate the conservation of charge.

@@

All fluorescence microscopy techniques that use an \textit{optical point spread
  function} or an \textit{optical transfer function} to model the mapping
between the fluorophores and the irradiance on the detector implicitly make the
monopole approximation. The optical point spread function is the irradiance
response of an optical system to an isotropic point source, so it cannot model
the response due to an anisotropic dipole radiator. In this work we define
\textit{monopole and dipole transfer functions} that describe the mapping
between fluorophores and the measured irradiance. Although optical systems are
an essential part of microscopes, fluorescence microscopists are interested in
measuring the properties of fluorophores (not optics), so the monopole and
dipole transfer functions are more directly useful than the optical transfer
function for the problems that fluorescence microscopists are interested in
solving.

@@

This work lies at the intersection of three fluorescence microscopy communities:
(1) the spatial ensemble imaging community where many fluorophores are imaged in
parallel and the goal is to find their concentration throughout the sample, (2)
the spatio-angular ensemble community where many fluorophores are imaged in
parallel and the goal is to find their concentration and orientation, and (3)
the SMLM community where fluorophores are sparse in the sample and the goal is
to find the position of each fluorophore. We briefly review how these three
communities use the monopole and scalar approximations.

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and all existing restoration techniques make either or both of the
monopole and scalar approximations. As discussed above, the Gibson-Lanni model,
the Born-Wolf model, and the Richards-Wolf model are approximate when applied to
fluorescence microscopes because they only model monopole emitters.
Deconvolution algorithms that use these models may make biased estimates of
fluorophore concentrations since they ignore the dipole excitation and emission
of fluorophores, but we will see that these effects are small when the sample
contains many randomly oriented fluorophores within a resolvable volume or
unconstrained rotating fluorophores.

@@

The most precise experiments in the SMLM community do not adopt the scalar or
monopole approximations. Although many works have applied dipole models with
vector optics, fewer have considered the effects of rotational or spatial
diffusion, and to our knowledge no works have considered both rotational and
spatial diffusion together. We will see that the dipole transfer functions are
useful tools for incorporating angular and spatial diffusion into SMLM
simulations and reconstructions.

In this work we place these three communities on a common theoretical footing.
We begin in section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In section \ref{sec:results} we use the mathematical
tools we developed in section \ref{sec:theory} to analyze and simulate a
single-view fluorescence microscope, and we demonstrate our techniques with four
simple phantoms. Finally, in section \ref{sec:discussion} we discuss the results
and their broader implications.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to unify the analysis of spatio-angular fluorescence microscopes.
In section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping between
object and data space in two different bases. In section \ref{sec:dipole} we
will extend the monopole imaging model to dipoles and examine the mapping in
four different bases.

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-transfer-functions/monopole-transfer-functions.pdf}
  \caption{The incoherent monopole transfer functions are related by a
    two-dimensional Fourier transform (right column). The coherent monopole
    transfer functions (left column) can be used to simplify the calculation of
    the remaining transfer functions.}
   \label{fig:monopole-transfer-functions}
 \end{figure}

@@

\begin{figure}
  \hspace{-2em}
  \includegraphics[scale=1.0]{../figures/transfer-functions/transfer-functions.pdf}
  \caption{There is one incoherent transfer function for each set of
    object-space basis functions, and these transfer functions are related by
    two-dimensional and spherical Fourier transforms---see center and right
    columns. There is an additional pair of coherent transfer functions that
    are useful for calculating the incoherent transfer functions---see left column.}
   \label{fig:transfer-functions}
 \end{figure}

@@

\bibliography{paper}
\appendix
\section{Spherical harmonics and the spherical Fourier transform}\label{sec:sph}
The spherical harmonic function of degree $\ell$ and order $-\ell \leq m \leq \ell$
is defined as \cite{schaeffer2013}
\begin{align}
Y_\ell^m(\vartheta, \varphi) = \sqrt{\frac{2\ell+1}{4\pi}}\sqrt{\frac{(\ell-|m|)!}{(\ell+|m|)!}}P_\ell^m\left(\cos\vartheta\right)\exp(i m \varphi),
\end{align}
where $P_\ell^m(\cos\theta)$ are the associated Legendre polynomials with the
Condon-Shortley phase
\begin{align}
  P_\ell^m(x) = (-1)^m(1-x^2)^{|m|/2}\frac{d^{|m|}}{dx^{|m|}}P_\ell(x),
\end{align}
and $P_\ell(x)$ are the Legendre polynomials defined by the recurrence
\begin{align}
  P_0(x) &= 1,\\
  P_1(x) &= x,\\
  \ell P_\ell(x) &= (2\ell-1)xP_{\ell-1}(x) - (\ell-1)P_{\ell-2}(x). 
\end{align}
The spherical harmonics are orthonormal, which means that
\begin{align}
  \int_{\mbb{S}^2}d\mh{s}\, Y_\ell^m(\mh{s}){Y}_{\ell'}^{m'*}(\mh{s}) = \delta_{\ell\ell'}\delta_{mm'},
\end{align}
where $\delta_{\ell\ell'}$ denotes the Kronecker delta. The spherical harmonics form a
complete basis, so an arbitrary function on the sphere $f(\mh{s})$ can be
expanded into a sum of weighted spherical harmonic functions
\begin{align}
  f(\mh{s}) = \sum_{\ell=0}^{\infty}\sum_{m=-\ell}^{l}F_\ell^mY_\ell^m(\mh{s}).
\end{align}
We can find the spherical harmonic coefficients $F_\ell^m$ for a given function
using Fourier's trick---multiply both sides by $Y_\ell^{m*}(\mh{s})$,
integrate over the sphere, and exploit orthogonality to find that
\begin{align}
  F_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, f(\mh{s})Y_\ell^{m*}(\mh{s}).
\end{align}
The coefficients $F_\ell^m$ are called the \textit{spherical Fourier transform}
of a spherical function.

@@

For $n=0$:
\begin{align}
  \text{chat}_0(\nu) &= 4\left[\int_0^{1/2}\tau d\tau\int_0^{\cos^{-1}\nu}d\phi_{\tau} - \int_{0}^{\nu/2}d\tau_x\int_0^{\frac{\tau_x}{\nu}\sqrt{1 - \nu^2}}d\tau_y\right]\Pi\left(\frac{\nu}{2}\right),\\
  \text{chat}_0(\nu) &= \frac{1}{2}\left[\cos^{-1}|\nu| - |\nu|\sqrt{1 - \nu^2}\right]\Pi\left(\frac{\nu}{2}\right),
\intertext{which is a well-known result \cite{goodman1996, mertz2009, bracewell2004}. For $n=1$:}
  \text{chat}_1(\nu) &= 16\Bigg[\int_0^{1/2}\tau d\tau\int_0^{\cos^{-1}\nu}d\phi_{\tau}(\tau^2 - \nu\tau\cos\phi_{\tau}) - \\ &\qquad\quad\int_{0}^{\nu/2}d\tau_x\int_0^{\frac{\tau_x}{\nu}\sqrt{1 - \nu^2}}d\tau_y\, (\tau_x^2 + \tau_y^2 - \nu\tau_x)\Bigg]\Pi\left(\frac{\nu}{2}\right),\\
  \text{chat}_1(\nu) &= \frac{1}{2}\left[\cos^{-1}|\nu| - |\nu| (3 - 2\nu^2)\sqrt{1 - \nu^2}\right]\Pi\left(\frac{\nu}{2}\right). 
\end{align}

>>>>

\section{Introduction}
Fluorescence microscopes are widely used in the biological sciences for
measuring ensembles of fluorescent molecules that label specific proteins and
biologically important molecules. While most fluorescence microscopy experiments
are designed to measure only the spatial distribution of fluorophores, a growing
number of experiments seek to measure both the spatial and angular distributions
of fluorophores by the use of polarizers \cite{vrabioiu2006, mattheyses2010,
  mehta2016, mcquilken2017, zhanghao2017} or point spread function engineering
\cite{agrawal2012, zhang2018}.

@@

We use the term \textit{monopole} to refer to a model of a fluorophore that
treats it as an isotropic absorber/emitter. Although the term \textit{monopole
  approximation} is not in widespread use, we think it accurately describes the
way many models of fluorescence microscopes treat fluorophores, and we use the
term to distinguish the monopole model from more realistic dipole and
higher-order models. Despite their use in models, electromagnetic monopole
absorber/emitters do not exist in nature. All physical fluorophores absorb and
emit radiation with dipole or higher-order moments, and these moments are always
oriented in space. For a classical mental model of fluorophores we imagine each
dipole as a small oriented antenna (with incoherent absorption and emission
moments) where electrons are constrained to move along a single direction. With
this model we can argue classically that monopole radiators do not exist because
they would violate the conservation of charge.

@@

All fluorescence microscopy models that use an \textit{optical point spread
  function} or an \textit{optical transfer function} to describe the mapping
between the fluorophores and the irradiance on the detector implicitly make the
monopole approximation. The optical point spread function is the irradiance
response of an optical system to an isotropic point source, so it cannot model
the response due to an anisotropic dipole radiator. In this work we define
\textit{monopole and dipole transfer functions} that describe the mapping
between fluorophores and the measured irradiance. Although optical systems are
an essential part of microscopes, fluorescence microscopists are interested in
measuring the properties of fluorophores (not optics), so the monopole and
dipole transfer functions are more directly useful than the optical transfer
function for the problems that fluorescence microscopists are interested in
solving.

@@

This work lies at the intersection of three subfields of fluorescence
microscopy: (1) spatial ensemble imaging where fluorophores are imaged as
ensembles and the goal is to find their concentration as a function of position
in the sample, (2) the spatio-angular ensemble imaging where fluorophores are
imaged as ensembles and the goal is to find their concentration and orientation
as a function of position in the sample, and (3) SMLM imaging where fluorophores
are sparse in the sample and the goal is to find the position and orientation of
each fluorophore. We briefly review how these three subfields use the monopole
and scalar approximations.

The large majority of fluorescence microscopes are used to image ensembles of
fluorophores, and all existing restoration techniques make either or both of the
monopole and scalar approximations. As discussed above, the Gibson-Lanni model,
the Born-Wolf model, and the Richards-Wolf model are approximate when applied to
fluorescence microscopy data because they only model monopole emitters.
Deconvolution algorithms that use these models may make biased estimates of
fluorophore concentrations since they ignore the dipole excitation and emission
of fluorophores, but we will see that these effects are small when the sample
contains many randomly oriented fluorophores within a resolvable volume or
unconstrained rotating fluorophores.

@@

The most precise experiments in SMLM imaging do not adopt the scalar or monopole
approximations. Although many works have applied dipole models with vector
optics, fewer have considered the effects of rotational or spatial diffusion,
and to our knowledge no works have considered both rotational and spatial
diffusion together. We will see that the dipole transfer functions are useful
tools for incorporating angular and spatial diffusion into SMLM simulations and
reconstructions.

In this work we place these three subfields on a common theoretical footing. We
begin in section \ref{sec:theory} by formulating the imaging problem
mathematically. We start with an abstract description of fluorescence imaging
systems and extend the usual monopole imaging model to dipoles. Along the way we
introduce transfer functions that will allow us to efficiently analyze
spatio-angular microscopes. In section \ref{sec:results} we use the mathematical
tools we developed in section \ref{sec:theory} to analyze and simulate a
single-view fluorescence microscope, and we demonstrate our techniques with four
simple phantoms. Finally, in section \ref{sec:discussion} we discuss the results
and their broader implications.

@@

The above discussion is quite abstract, but it is a powerful point of view that
will enable us to unify the analysis of spatio-angular fluorescence imaging.
In section \ref{sec:monopole} we will demonstrate the formalism by examining a
familiar monopole imaging model, and we will demonstrate the mapping between
object and data space in two different bases. In section \ref{sec:dipole} we
will extend the monopole imaging model to dipoles and examine the mapping in
four different bases.

@@

\begin{figure}
  \centering
  \includegraphics[scale=1.0]{../figures/monopole-transfer-functions/monopole-transfer-functions.pdf}
  \caption{The monopole transfer functions are related by a
    two-dimensional Fourier transform (right column). The coherent monopole
    transfer functions (left column) can be used to simplify the calculation of
    the remaining transfer functions.}
   \label{fig:monopole-transfer-functions}
 \end{figure}

@@

\begin{figure}
  \hspace{-2em}
  \includegraphics[scale=1.0]{../figures/transfer-functions/transfer-functions.pdf}
  \caption{There is one transfer function for each set of object-space basis
    functions, and these transfer functions are related by two-dimensional and
    spherical Fourier transforms---see center and right columns. There is an
    additional pair of coherent transfer functions that are useful for
    calculating the transfer functions---see left column.}
   \label{fig:transfer-functions}
 \end{figure}

@@

\bibliography{paper}
\appendix
\section{Spherical harmonics and the spherical Fourier transform}\label{sec:sph}
The spherical harmonic function of degree $\ell$ and order $-\ell \leq m \leq \ell$
is defined as \cite{schaeffer2013}
\begin{align}
Y_\ell^m(\vartheta, \varphi) = \sqrt{\frac{2\ell+1}{4\pi}}\sqrt{\frac{(\ell-|m|)!}{(\ell+|m|)!}}P_\ell^m\left(\cos\vartheta\right)\exp(i m \varphi),
\end{align}
where $P_\ell^m(\cos\vartheta)$ are the associated Legendre polynomials with the
Condon-Shortley phase
\begin{align}
  P_\ell^m(x) = (-1)^m(1-x^2)^{|m|/2}\frac{d^{|m|}}{dx^{|m|}}P_\ell(x),
\end{align}
and $P_\ell(x)$ are the Legendre polynomials defined by the recurrence
\begin{align}
  P_0(x) &= 1,\\
  P_1(x) &= x,\\
  \ell P_\ell(x) &= (2\ell-1)xP_{\ell-1}(x) - (\ell-1)P_{\ell-2}(x). 
\end{align}
The spherical harmonics are orthonormal, which means that
\begin{align}
  \int_{\mbb{S}^2}d\mh{s}\, Y_\ell^m(\mh{s}){Y}_{\ell'}^{m'*}(\mh{s}) = \delta_{\ell\ell'}\delta_{mm'},
\end{align}
where $\delta_{\ell\ell'}$ denotes the Kronecker delta. The spherical harmonics form a
complete basis, so an arbitrary function on the sphere $f(\mh{s})$ can be
expanded into a sum of weighted spherical harmonic functions
\begin{align}
  f(\mh{s}) = \sum_{\ell=0}^{\infty}\sum_{m=-\ell}^{l}F_\ell^mY_\ell^m(\mh{s}).
\end{align}
We can find the spherical harmonic coefficients $F_\ell^m$ for a given function
using Fourier's trick---multiply both sides by $Y_\ell^{m*}(\mh{s})$,
integrate over the sphere, and exploit orthogonality to find that
\begin{align}
  F_\ell^m = \int_{\mbb{S}^2}d\mh{s}\, f(\mh{s})Y_\ell^{m*}(\mh{s}).
\end{align}
The coefficients $F_\ell^m$ are called the \textit{spherical Fourier transform}
of a spherical function.

@@

For $n=0$:
\begin{align}
  \text{chat}_0(\nu) &= 4\left[\int_0^{1/2}\tau d\tau\int_0^{\cos^{-1}\nu}d\phi_{\tau} - \int_{0}^{\nu/2}d\tau_x\int_0^{\frac{\tau_x}{\nu}\sqrt{1 - \nu^2}}d\tau_y\right]\Pi\left(\frac{\nu}{2}\right),\\
  \text{chat}_0(\nu) &= \frac{1}{2}\left[\cos^{-1}|\nu| - |\nu|\sqrt{1 - \nu^2}\right]\Pi\left(\frac{\nu}{2}\right),
\intertext{which is a well-known result \cite{goodman1996, mertz2009, bracewell2004}. For $n=1$:}
  \text{chat}_1(\nu) &= 16\Bigg[\int_0^{1/2}\tau d\tau\int_0^{\cos^{-1}\nu}d\phi_{\tau}(\tau^2 - \nu\tau\cos\phi_{\tau}) - \nonumber\\ &\qquad\quad\int_{0}^{\nu/2}d\tau_x\int_0^{\frac{\tau_x}{\nu}\sqrt{1 - \nu^2}}d\tau_y\, (\tau_x^2 + \tau_y^2 - \nu\tau_x)\Bigg]\Pi\left(\frac{\nu}{2}\right),\\
  \text{chat}_1(\nu) &= \frac{1}{2}\left[\cos^{-1}|\nu| - |\nu| (3 - 2\nu^2)\sqrt{1 - \nu^2}\right]\Pi\left(\frac{\nu}{2}\right). 
\end{align}
